Mendelian randomization for causal inference in genome-wide association studies

### 2022-03-01 mr-review.Rmd
organized bidirectional papers into one section
added text for MR TRYX

@rees_robust_2019 approached the challenge of horizontal pleiotropy by 
developing three new methods for two-sample MR analysis with GWAS summary data: robust 
regression, penalized weights, and LASSO penalization. The first two can be viewed as 
modifications of MR-Egger and IVW methods. Together, they offer three strategies for 
downweighting or excluding variants with heterogeneous causal estimates.
Recall that MR-Egger offers consistent causal effect estimates when there is no 
horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the 
"Instrument Strength Independent of Direct Effect" (InSIDE) assumption [@bowden_mendelian_2015]. The 
InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP 
instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses 
MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to 
formulate an estimator with high breakdown point and high efficiency. The second estimation 
method, with penalized weights, is much like the penalized median estimator, from above, 
except that the second argument in the minimum function is $100q_j$, instead of the $20q_j$ 
above. Finally the LASSO-based method fashions an estimator by applying a penalty to the 
objective function from MR-Egger to get Equation \@ref(eq:lasso) as the needed objective function.

@@

@wang_causal_2021 developed a framework, GRAPPLE, for performing two-sample MR with weak and strong 
instruments. Using GWAS summary statistics as inputs, GRAPPLE extends MR-RAPS and can detect presence of horizontally 
pleiotropic pathways, infer the causal direction, and perform multivariable MR. Central to the 
GRAPPLE framework is the observation that a horizontally pleiotropic pathway often gives rise to an 
additional mode in the profile likelihood. GRAPPLE uses the presence of multiple profile likelihood 
modes to diagnose horizontal pleiotropy with effects that are grouped by pathway. 
GRAPPLE facilitates identification of SNP instruments that contribute to the 
additional profile likelihood modes. A researcher using GRAPPLE may then recognize 
a confounding factor for each mode. With additional GWAS summary data with traits 
that reflect the confounding factors, GRAPPLE can fit multivariable MR models when
the InSIDE assumption holds for the remaining horizontally pleiotropic effects [@wang_causal_2021].

@@

@darrous_simultaneous_2021

>>>>

@rees_robust_2019 approached the challenge of horizontal pleiotropy by 
developing three new methods for two-sample MR analysis with GWAS summary data: robust 
regression, penalized weights, and LASSO penalization. The first two can be viewed as 
modifications of MR-Egger and IVW methods. Together, they offer three strategies for 
downweighting or excluding variants with heterogeneous causal estimates.
Recall that MR-Egger offers consistent causal effect estimates when there is no 
horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the 
"Instrument Strength Independent of Direct Effect" (InSIDE) assumption [@bowden_mendelian_2015]. The 
InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP 
instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses 
MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to 
formulate an estimator with high breakdown point and high efficiency. The second estimation 
method, with penalized weights, is much like the penalized median estimator, from above, 
except that the second argument in the minimum function is $100q_j$, instead of the $20q_j$ 
above. Finally the LASSO-based method fashions an estimator by applying a penalty to the 
objective function from MR-Egger to get Equation \@ref(eq:lasso) as the objective function.

@@

@cho_exploiting_2020 developed the MR-TRYX framework for exploiting horizontal pleiotropy
to identify alternative causal pathways. They did this by positing that, for a 
collection of putative SNP instruments, any heterogeneity in the Wald ratio estimates
may be due to horizontal pleiotropy. Once they identify a SNP with outlying 
ratio estimates, they search GWAS results by querying the MR-Base database to identify complex traits (in other studies) that 
associate with the outlying SNP [@hemani_mr-base_2018]. They follow up promising trait associations to 
account for their observations with multiple causal pathways.

@@

@darrous_simultaneous_2021
@wang_causal_2021 developed a framework, GRAPPLE, for performing two-sample MR with weak and strong 
instruments. Using GWAS summary statistics as inputs, GRAPPLE extends MR-RAPS and can detect presence of horizontally 
pleiotropic pathways, infer the causal direction, and perform multivariable MR. Central to the 
GRAPPLE framework is the observation that a horizontally pleiotropic pathway often gives rise to an 
additional mode in the profile likelihood. GRAPPLE uses the presence of multiple profile likelihood 
modes to diagnose horizontal pleiotropy with effects that are grouped by pathway. 
GRAPPLE facilitates identification of SNP instruments that contribute to the 
additional profile likelihood modes. A researcher using GRAPPLE may then recognize 
a confounding factor for each mode. With additional GWAS summary data with traits 
that reflect the confounding factors, GRAPPLE can fit multivariable MR models when
the InSIDE assumption holds for the remaining horizontally pleiotropic effects [@wang_causal_2021].
